MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-04 main/MSstatsPTM_Main.tex
made adjustments to supp and main

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The model measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein. We implement this model in the free and open-source R package $MSstatsPTM$.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward. \cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.

@@

The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and limma methods. The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT. \cite{Choi:2014} \cite{Huang:2020}

@@

The proposed statistical approach methods were evaluated using two computer simulations, with known ground truth. Specifically, their properties under adjustment with respect to protein abundance were evaluated. Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}.

@@

In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values. Multiple simulations were generated with these parameters, with different numbers of replicates and conditions. 1000 modified peptides were generated, half of which were true positives and the other half true negatives. Of true positives simulated, half of them had their abundance's masked by changes in global protein. Modified peptides and unmodified proteins were both simulated with 10 features.

@@

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. Missing values and feature count values for this simulation were determined by looking at the averages from the biological experiments in this paper. Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were masked completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.

@@

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

@@

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously. In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantial less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs. Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

@@

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package will only summarize features, PSMs, which include the same modification at the same site. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. As mentioned in Section \ref{sec:missing_imp}, the package uses an AFT model to impute missing values in this step, although this step is optional.

The final step of the package is modeling the summarized dataset. A linear mixed effects model is automatically fit for both the summarized modified and global profiling runs. This model is automatically adjusted depending on the experimental design and acquisition method. The comparisons of interest can either be predefined or a full pairwise comparison will be tested. After modeling the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.

Beyond the core conversion, summary, and modeling functionality, the package also includes two plotting functions. The first creates plots for the summarization data to help with quality assurance analysis and identifying sources of variation. This function plots a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and profile plot, plotting each feature and the overall feature summarization as a line plot over each run. The second function plots the modeling results and includes a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.

The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, depending on the data acquisition type. The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.  All datasets in this paper were evaluated using the $MSstatsPTM$ package.

@@

The performance of of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)). For biological experiments where ground truth was not known, we evaluated significantly differential peptides with an $\alpha$ of .05. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.

@@

In this simulation, not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and limma performed similarly, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.

@@

Changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and outperformed the other methods. Additionally, in the recall rate the proposed method showed the strongest performance, even when the number of biological replicates was low. Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions. Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.

@@

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is more in line with expectation. Additionally, background peptides serving as the null model, colored grey, show many false positives before adjustment is made. Again after adjustment the results improve and the number of false positives decrease significantly.

In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, using limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.

@@

Looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect. The results of this can be seen in Figure \ref{fig:ipah_profile}.

@@

Two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.  The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.015$ between the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$. In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected. This is contrast with a negative fold change in the global profiling run of $-.616$. In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$. The profile plots of these two PTMs can be seen in Figure \ref{fig:shigella_profile}

@@

As discussed previously, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted. Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.

Given the lack of matches in the global profiling run, the proposed method was unable to be applied to as many modified peptides as in the previous examples. However we can still see the power of the method when looking at peptide $P52209\_K059$. The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein. Specifically looking at the comparison between the CCCP and Control conditions, the unadjusted model showed a small log fold change of $-.255$ and insignificant adjusted pvalue of $.104$. However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$. The profile plot for this PTM can be seen in Figure \ref{fig:usp30_profile}. Using the proposed method we get increased information that would otherwise have been missed.

@@

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig1.png}
\caption{\todo{todo}}
\label{fig:workflow}
\end{figure}

>>>>

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The method measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein. We implement this model in the free and open-source R package $MSstatsPTM$.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward.\cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.

@@

The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and Limma methods.\cite{Ritchie_15a} The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT.\cite{Choi:2014} \cite{Huang:2020}

@@

The proposed statistical approach methods were evaluated using two computer simulations, with known ground truth. Specifically, their properties under adjustment with respect to protein abundance were evaluated. Two simulations were ran, mimicking a label free experiment, one including a high number of features and no missing values, and one with few modified features and including missing values.

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or Limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and Limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}.

@@

Here simulations were generated with a high number of modified and unmodified peptide features and no missing values. Multiple simulations were generated with different numbers of replicates and conditions. In all simulations 1000 modified peptides were generated, half of which were true positives and the other half true negatives. Half of true positives and true negatives were generated with convolution with global protein abundance. Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential. Modified peptides and unmodified proteins were both simulated with 10 features.

@@

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. The percentage of missing values and feature counts were determined by looking at the averages from the biological experiments in this paper. Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were masked completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance. Otherwise this simulation used the same parameters as in simulation 1, with 1000 modified peptides half of which were positives and the other half negative.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The four mixtures were compared with known fold changes for the spike-in peptides.

@@

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

@@

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously. In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantial less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides. In this case there only 41.9\% of the modified peptides had a corresponding unmodified protein to perform adjustment. In cases where there is no estimate for the matching unmodified protein, adjustment cannot be performed.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.

Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

@@

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features, PSMs, which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package uses an AFT model, as mentioned in Section \ref{sec:missing_imp}, to impute missing values, although this step is optional.

The final step of the package is modeling the summarized dataset. A linear mixed effects model is automatically fit for both the summarized modified and global profiling runs. This model is automatically adjusted depending on the experimental design and acquisition method. The comparisons of interest can either be predefined or a full pairwise comparison will be tested. After fitting a model to both the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.

Beyond the core functionalities of conversion, summary, and modeling, the package also includes functions for plotting the results. These include plots for the summarized and modeled data to assist in the analysis of the experiment. The summarized plots help with quality assurance analysis and identifying sources of variation. This includes a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and a profile plot, plotting each feature and the overall feature summarization as a line plot over each run. Additionally, the model plots include a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.

The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, depending on the data acquisition type. The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)). For biological experiments where ground truth was not known, we evaluated significantly differential peptides with an $\alpha$ of .05. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

@@

Here we see that not accounting for changes in unmodified peptide level results in a high false discovery rate, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.

@@

In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and the recall rate, even when the number of replicates were low. Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. In comparison, Limma and t-test showed much wider distributions. This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}. Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.

@@

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is more in line with expectation. Additionally, the background peptides serving as the null model, colored grey, show many false positives before adjustment is made. Again after adjustment the results improve and the number of false positives decrease significantly.

In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using Limma with protein adjustment. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using Limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, using Limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.

@@

The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}. In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are correct in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

In Figure \ref{fig:ipah_profile}, looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect.

@@

The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}. In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 39,550 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance. This is still a very small portion (4.02\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

In Figure \ref{fig:shigella_profile} the profile plot of two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.  The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.015$ between the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$. In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected. This is contrast with a negative fold change in the global profiling run of $-.616$. In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.

@@

As discussed previously, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted. Additionally, not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.

The results of this experiment are summarized in Figures \ref{fig:usp30_ven_d} and \ref{fig:usp30_profile}. In \ref{fig:usp30_ven_d} the number of significant modified peptides before and after protein adjustment is shown. In the top plot we can see there are a very large number of peptides that become insignificant after adjustment. This is mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run. In the bottom plot only modified peptides that could be adjusted are shown. Here there are much less peptides that become insignificant after adjustment. In this plot we see that 726 peptides became insignificant, 547 became significant, and 1,078 are significant in both models. As in the previous experiments, we can check if these peptides are becoming insignificant due to an increase in variance. To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment. In this experiment there are only 25 peptides that meet this criteria, a very small percentage of all modified peptides.

In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown. The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein. Specifically looking at the comparison between the CCCP and Control conditions, the unadjusted model showed a small log fold change of $-.255$ and insignificant adjusted pvalue of $.104$. However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$. Using the proposed method we get increased information that would otherwise have been missed.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code availability}

\bf{Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html

\noindent{\bf{Github:} https://github.com/Vitek-Lab/MSstatsPTM}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig1.png}
\caption{\todo{Meena?: Write about how MS for PTMs is run?}}
\label{fig:workflow}
\end{figure}
